---
title: Creating new alabaster applications
author:
- name: Aaron Lun
  email: infinite.monkeys.with.keyboards@gmail.com
package: alabaster.base
date: "Revised: September 29, 2022"
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Creating new applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(BiocStyle)
self <- Githubpkg("ArtifactDB/alabaster.base");
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Overview

Developers can also create "alabaster applications", which customize the machinery of the **alabaster** framework for specific needs.
Customizations typically involve:

- Changing how files/metadata are retrieved when loading objects into an R session.
- Added new schemas to support application-specific classes.
- Adding more metadata requirements to some or existing classes.

We provide a number of hooks inside `r self` to enable applications to apply these customizations over the defaults.

# Creating a new retrieval mechanism

The `acquireMetadata()` and `acquireFile()` generic, which describe how the metadata and file artifacts should be acquired.
For example, applications can write new methods that query APIs for this content rather than reading it from the filesystem.

```{r}
library(alabaster.base)
setClass("RemoteDownloadHandler", slots=c(host="character", id="character"))

setMethod("acquireFile", "RemoteDownloadHandler", function(project, path) {
    # Download a file here, possibly with some application-specific caching.
    demo.url <- paste0(project@host, "/projects/", project@id, "/file", path)
    target <- tempfile()
    download.file(demo.url, target)
    target
})

setMethod("acquireMetadata", "RemoteDownloadHandler", function(project, path) {
    # Pull down JSON metadata, possibly with some application-specific caching.
    demo.url <- paste0(project@host, "/projects/", project@id, "/file", path)
    target <- tempfile()
    download.file(demo.url, target)
    jsonlite::fromJSON(target, simplifyVector=TRUE, simplifyMatrix=FALSE, simplifyDataFrame=FALSE)
})
```

We can then construct an instance of a `RemoteDownloadHandler`;

```{r}
proj <- new("RemoteDownloadHandler", host="https://my.institute.org/downloads", id="PROJECT_ID")
```

And use it in `loadObject()`, which will use the new methods for all internal `acquireMetadata()` and `acquireFile()` calls.
This will instruct `loadObject()` to download remote content on demand.

```{r, eval=FALSE}
# Demonstration only.
meta <- acquireMetadata(proj, "some/file/path.csv")
obj <- loadObject(meta, proj)
```

# Creating new schemas

Applications can define their own schemas with custom metadata requirements or data structures.
See [here](https://github.com/ArtifactDB/BiocObjectSchemas) for more details.
Once created, the application-specific schemas should be stored in the `inst/schemas` directory of the application's R package.

# Overriding the staging method

The `.altStageObject()` setter allows applications to replace `stageObject()` with an alternative staging generic.
This can be used to perform additional tasks during staging of some or all classes, e.g., add more metadata.
For example, we could force all saving operations to include the author in the metadata for non-child objects:

```{r}
setGeneric("appStageObject", function(x, dir, path, child=FALSE, ...) 
    standardGeneric("appStageObject"))

setMethod("appStageObject", "ANY", function(x, dir, path, child=FALSE, ...) {
    meta <- stageObject(x, dir, path, child=child, ...)

    # Creating a fallback that adds authorship to all non-child objects. 
    if (!child) {
        meta$authors <- I(Sys.info()[["user"]])
    }

    # Pointing .writeMetadata to the application package storing the schemas.
    attr(meta[["$schema"]], "package") <- "APPLICATION_NAME_HERE"

    meta 
})
```

Application developers will typically create another wrapper function that runs the setter before attempting to stage an object.
This is more convenient for the end-users of that application, who do not have to remember what to set:

```{r}
appSave <- function(x, dir, path, child=FALSE) {
    olds <- .altStageObject(appStageObject)
    on.exit(.altStageObject(olds))

    meta <- appStageObject(x, dir, path, child=child)
    written <- .writeMetadata(meta, dir) 

    written$path
}
```

Of course, it is possible to customize the staging of specific classes by simply defining a new method for the new generic:

```{r}
setMethod("appStageObject", "DFrame", function(x, dir, path, child=FALSE, ...) {
    # Extracting details from the metadata.
    clinical.trial.id <- metadata(x)$trial_id
    treatment <- metadata(x)$treatment
    study.design <- metadata(x)$study

    # Calling the ANY method, which eventually just calls stageObject...
    meta <- callNextMethod()

    # Adding some metadata for data frames, assuming these properties are
    # listed in the schema.
    meta[["clinical_trail_details"]] <- list(
        trial_id = trail.id,
        treatment = treatment,
        design = study.design
    )

    meta
})
```

# Overriding the loading method

Similarly, the `.altLoadObject()` setter allows applications to set an alternative loading function to replace `loadObject()`.
This is necessary if `.altStageObject()` is used to define a staging override that points to a different set of schemas;
in which case, the loading method must also be overridden to look at those schemas to obtain appropriate restoration method (in the `_attributes.restore.R` property).
The loading method may also attach various bits and pieces of global metadata that we might have stored, e.g. authorship.

```{r}
# Memorize schema look-up in the R session, to avoid having to repeatedly query
# the R package directory on the file system at every loadObject request.
memory <- new.env()
memory$cache <- list()

appLoadObject <- function(info, project, ...) {
    output <- .loadObjectInternal(info, project, ..., 
        locations="APPLCATION_NAME_HERE",
        memory=memory.env
    )

    if ("authors" %in% names(info) && is(output, "Annotated")) {
        metadata(output)$authors <- info$authors
    }

    output
}
```

Application developers will typically create another wrapper function that runs the setter before attempting to load an object.
This wrapper may also create the application-specific `*Handler` class that will be used to override `acquireFile` and `acquireMetadata`.

```{r}
appLoad <- function(project, path) {
    oldl <- .altLoadObject(appLoadObject)
    on.exit(.altLoadObject(oldl))
    info <- acquireMetadata(project, path)
    appLoadObject(info, project, child=child)
}
```

Note that `.altLoadObject()` will override the loading for all objects.
For overrides of specific objects, application developers can simply modify the `_attributes.restore.R` in the corresponding schema to point to a different function.

# Session information {-}

```{r}
sessionInfo()
```
